---
title: "Understanding Afterschool Programs in Dallas County"
output:
  distill::distill_article:
    toc: TRUE
    theme: theme.css
---

To understand the landscape of afterschool programs across Dallas County we assessed two components. The existing afterschool environment and current neighborhood characteristics. Comparing these components  will help identify areas in need of additional afterschool programming.

```{r, map of DAS afterschool programs in county, echo = FALSE, message=FALSE, warning=FALSE, layout="l-body", fig.height=8}
# Generate Jenks Breaks for plot
brks_das <- BAMMtools::getJenksBreaks(dasmetrics$ProgSqMi, k = 6)
dasmetrics$group <- cut(dasmetrics$ProgSqMi,
                             breaks = brks_das,
                       include.lowest = TRUE)
labs_das <- round(brks_das)

# Generate labels based on Jenks breaks for plot
labs_plot <- paste0("(", labs_das[1:5], " - ", labs_das[2:6], ")")

# Generate Plot
ggplot() +
  geom_sf(data = dasmetrics, aes(fill = group), color = "#595959") +
  geom_sf(data = dallasCounty, color = "#323232", fill = NA, size = 2) +
  geom_sf(data = highways, color = "#323232", size = 1) +
  scale_fill_brewer(name = "",
                    palette = "BuPu",
                    label = labs_plot,
                    guide = guide_legend(direction = "horizontal",
                                         nrow = 1,
                                         label.position = "top")) +
  labs(title = "After School Program Density (per Square Mile)",
       #subtitle = "U.S. Census Bureau, American Community Survey 5-Year Estimates 2020",
       x = "",
       y = "",
       color = "") +
  theme_void() +
  theme(legend.position = "bottom",
        legend.key.width = unit(0.5, units = "in"),
        plot.caption = element_text(size = 36, face = "italic"),
        legend.spacing = unit(0, "in"), 
        legend.key.height = unit(0.25, "in"), 
        legend.spacing.x = unit(0.1, "in"), 
        legend.justification = "center",
        legend.margin=margin(4,4,4,4),
        legend.box.spacing = margin(0.5))

```

## Existing Afterschool Environment

Using program data from Dallas Afterschool, we developed a measure of the existing afterschool environment (supply) by comparing afterschool program density, and afterschool eligible children to total seats, and program count. 

### Summary of Current Programs

In Dallas County, programs affiliated with Dallas Afterschool predominantly use a sliding scale payment model and make up % percent of all programs. With sliding scale programs, children only need to pay an affordable fee relative to their household income in order to participate.

Additional facility types and fee structures are broken down below. 

```{r, table of programs, echo = FALSE, message=FALSE, warning=FALSE, layout="l-body-outset", fig.height=8}
das <- das_programs %>%
  st_drop_geometry(.) %>%
  group_by(facility_type) %>%
  dplyr::summarise(count = n(),
                   capacity = sum(max_capacity)) %>%
  mutate(#source = "Dallas Afterschool",
         avgseats = round(capacity/count),
         percap = capacity/sum(capacity),
         percapStr = paste0(round(percap*100, digits = 1), "%"),
         facility_type = ifelse(facility_type == "", "Not Available", facility_type))

#hhs <- hhs_programs %>%
#  st_drop_geometry(.) %>%
#  dplyr::summarise(count = n(),
#                   capacity = sum(capacity)) %>%
#  mutate(source = "Texas Health and Human Services",
#         avgseats = round(capacity/count))
#
#cacfp <- cacfp_programs %>%
#  st_drop_geometry(.) %>%
#  dplyr::summarise(count = n(),
#                   capacity = sum(site_license_capacity, na.rm = TRUE)) %>%
#  mutate(source = "Child and Adult Food Care Program",
#         avgseats = round(capacity/count))

das %>%
  select(-percap) %>%
  knitr::kable(.,
               col.names = c('Facility Type', 'Total Programs', 'Seat Capacity', 'Average Seats', 'Percent of Seats')) %>%
  kable_styling()
```

```{r, Supply Scores Histogram, echo = FALSE, message=FALSE, warning=FALSE}
das %>%
  ggplot(aes(x = percap, y = reorder(facility_type, -percap), fill = facility_type)) +
  geom_bar(stat = "identity", alpha = 0.7) +
  scale_x_continuous(labels = scales::percent) +
  scale_color_manual(values = CPALtools::pal_cpalfac) +
  scale_fill_manual(values = CPALtools::pal_cpalfac) +
  labs(title = "Percentage of Total Seats by Facility Type",
       #subtitle = "U.S. Census Bureau, American Community Survey 5-Year Estimates 2020",
       x = "",
       y = "",
       color = "") +
  theme_bw() +
  theme(legend.position = "none",
        legend.key.width = unit(0.5, units = "in"),
        plot.caption = element_text(size = 36, face = "italic"),
        legend.spacing = unit(0, "in"), 
        legend.key.height = unit(0.25, "in"), 
        legend.spacing.x = unit(0.1, "in"), 
        legend.justification = "center",
        legend.margin=margin(4,4,4,4),
        legend.box.spacing = margin(0.5))
```

### Programs in Dallas County

```{r, map of programs in dallas county, echo = FALSE, message=FALSE, warning=FALSE, layout="l-body", fig.height=8}
#unique(das_programs$facility_type)
ccc <- das_programs %>%
  filter(facility_type == "Child Care Center")
school <- das_programs %>%
  filter(facility_type == "School")
missing <- das_programs %>%
  filter(facility_type == "")
comcen <- das_programs %>%
  filter(facility_type == "Community Center/Nonprofit")
faith <- das_programs %>%
  filter(facility_type == "Faith Community")
munbui <- das_programs %>%
  filter(facility_type == "Municipal Building")
aptcom <- das_programs %>%
  filter(facility_type == "Apartment Complex")

leaflet() %>%
#  setView(lng = -96.7970, lat = 32.7767, zoom = 10) %>%
  addTiles(urlTemplate = cpal_mapbox, attribution = cpal_leaflet) %>%
    addCircleMarkers(data = ccc,
                     stroke = FALSE,
                     radius = 4,
                     fillColor = "#0089C4",
                     weight = 1,
                     fillOpacity = 1,
                     group = "Child Care Center"
              ) %>%
    addCircleMarkers(data = school,
                     stroke = FALSE,
                     radius = 4,
                     fillColor = "#DC5034",
                     weight = 1,
                     fillOpacity = 1,
                     group = "School"
              ) %>%
    addCircleMarkers(data = missing,
                     stroke = FALSE,
                     radius = 4,
                     fillColor = "#54494B",
                     weight = 1,
                     fillOpacity = 1,
                     group = "Not Available"
              ) %>%
    addCircleMarkers(data = comcen,
                     stroke = FALSE,
                     radius = 4,
                     fillColor = "#FFB533",
                     weight = 1,
                     fillOpacity = 1,
                     group = "Non-Profit"
              ) %>%
    addCircleMarkers(data = faith,
                     stroke = FALSE,
                     radius = 4,
                     fillColor = "#87F1FF",
                     weight = 1,
                     fillOpacity = 1,
                     group = "Faith Community"
              ) %>%
    addCircleMarkers(data = munbui,
                     stroke = FALSE,
                     radius = 4,
                     fillColor = "#ADB533",
                     weight = 1,
                     fillOpacity = 1,
                     group = "Municipal Building"
              ) %>%
    addCircleMarkers(data = aptcom,
                     stroke = FALSE,
                     radius = 4,
                     fillColor = "#8078B6",
                     weight = 1,
                     fillOpacity = 1,
                     group = "Apartment Complex"
              ) %>%
  addPolygons(data = dallasCounty,
              color = "#686158",
              weight = 3,
              smoothFactor = 0.2, 
              fillOpacity = 0,
              group = "Dallas County"
              ) %>%
  addLayersControl(
    baseGroups = c("Dallas County"),
    overlayGroups = c("Apartment Complex", "Child Care Center", "Faith Community", "Municipal Building", "Non-Profit", "School", "Not Available"),
    position = "topright",
    options = layersControlOptions(collapsed = FALSE))

```

## Current Neighborhood Conditions

We understand that the need for afterschool programs in any given neighborhood  will vary based on the demographic characteristics of local  families and children. To assess demand,  we identified a number of characteristics for residents who may have a higher need for afterschool programming. 

These characteristics include children experiencing  food insecurity, , parents who travel longer distances to  work or have irregular work schedules,, single parent households, and households facing unemployment. These are all considerations we took  into account to assess current neighborhood conditions and demand for afterschool programming. 


### Summary of Current Conditions

```{r, Eligible Population, echo = FALSE, message=FALSE, warning=FALSE}
dasmetrics %>%
  ggplot(aes(x = eligPop)) +
  geom_histogram(bins = 50, color = "#DC5034", fill = "#DC5034", alpha = 0.7) +
  labs(title = "Distribution of Eligible Population (5 - 14)",
       #subtitle = "U.S. Census Bureau, American Community Survey 5-Year Estimates 2020",
       x = "",
       y = "",
       color = "") +
  theme_bw() +
  theme(legend.position = "bottom",
        legend.key.width = unit(0.5, units = "in"),
        plot.caption = element_text(size = 36, face = "italic"),
        legend.spacing = unit(0, "in"), 
        legend.key.height = unit(0.25, "in"), 
        legend.spacing.x = unit(0.1, "in"), 
        legend.justification = "center",
        legend.margin=margin(4,4,4,4),
        legend.box.spacing = margin(0.5))
```

[Summary of population variables for Dallas County. How many families/children do we expect could go to Afterschool programs in the county?]
[Specific family types (Single-mother households, households below poverty, etc.)]