---
title: "Highest Need Afterschool Priority Areas"
output:
  distill::distill_article:
    toc: TRUE
    theme: theme.css
---

```{r Set-Up Block, include = FALSE}
# generate table by group
typologytable <- dasmetrics %>%
  st_drop_geometry(.) %>%
  group_by(MainGrp) %>%
  summarise(TotalTracts = n(),
            AreaSqMi = sum(AreaSqMi),
            mhi = median(med_incE, na.rm = TRUE),
            eligPop = sum(eligPop),
            bp_u18 = sum(bp_u18E, na.rm = TRUE),
            pop_u18 = sum(pop_u18E, na.rm = TRUE),
            dasPrograms = sum(dasPrograms),
            dasSeats = sum(dasSeats),
            #cacfpPrograms = sum(cacfpPrograms),
            #cacfpSeats = sum(cacfpSeats),
            #hhsPrograms = sum(hhsPrograms),
            #hhsSeats = sum(hhsSeats),
            #totPrograms = sum(dasPrograms+cacfpPrograms+hhsPrograms),
            #totSeats = sum(dasSeats+cacfpSeats+hhsSeats)
            ) %>%
  mutate(poppro = round(eligPop/dasSeats, digits = 3),
         cpr = round(bp_u18/pop_u18, digits = 3))

typeSupply <- dasmetrics %>%
  st_drop_geometry(.) %>%
  group_by(typeSupply) %>%
  dplyr::summarise(Supply = n()) %>%
  #mutate(typeSupply = str_remove(typeSupply, " SUPPLY")) %>%
  rename(Classification = typeSupply)

typeDemand <- dasmetrics %>%
  st_drop_geometry(.) %>%
  group_by(typeDemand) %>%
  dplyr::summarise(Demand = n()) %>%
  #mutate(typeDemand = str_remove(typeDemand, " DEMAND")) %>%
  rename(Classification = typeDemand)

typeJoin <- left_join(typeSupply, typeDemand) %>%
  mutate(Demand = ifelse(is.na(Demand), 0, Demand),
         Supply = ifelse(is.na(Supply), 0, Supply),
         Order = ifelse(Classification == "VERY HIGH", 1, 
                        ifelse(Classification == "HIGH", 2, 
                               ifelse(Classification == "MODERATE", 3, 
                                      ifelse(Classification == "LOW", 4, 
                                             ifelse(Classification == "VERY LOW", 5, "ERROR")))))) %>%
    mutate(Classification = ifelse(Classification == "VERY HIGH", "Very High", 
                        ifelse(Classification == "HIGH", "High", 
                               ifelse(Classification == "MODERATE", "Moderate", 
                                      ifelse(Classification == "LOW", "Low", 
                                             ifelse(Classification == "VERY LOW", "Very Low", "ERROR"))))))

unique(dasmetrics$MainGrp)
dem_sup <- dasmetrics %>%
  filter(MainGrp == "Demand > Supply")

sup_dem <- dasmetrics %>%
  filter(MainGrp == "Demand < Supply")

demTsup <- dasmetrics %>%
  filter(MainGrp == "Demand = Supply")

mean(dasmetrics$eligPop, na.rm = TRUE)
priorityareas <- dasmetrics %>%
  filter(brkdiff >= 2) %>%
  filter(mhiE <= mean(dasmetrics$mhiE, na.rm = TRUE)) %>%
  filter(eligPop >= mean(dasmetrics$eligPop, na.rm = TRUE)) %>%
  filter(cpr >= mean(dasmetrics$cpr, na.rm = TRUE)) %>%
  mutate(MainGrp = "Priority Area")
  
nopriorityareas <- dasmetrics %>%
  filter(brkdiff <= -2) %>%
  filter(mhiE > mean(dasmetrics$mhiE, na.rm = TRUE)) %>%
  filter(eligPop < mean(dasmetrics$eligPop, na.rm = TRUE)) %>%
  filter(cpr < mean(dasmetrics$cpr, na.rm = TRUE)) %>%
  mutate(MainGrp = "Non-Priority Area")
#plot(priorityareas["geometry"])
```

## Afterschool Program Need in Dallas County

The map below visualizes the gap between the supply and demand score in Dallas County. By toggling the layer selection on the top right of the map you will be able to view each score. By clicking on any given census tract a popup will appear as well displaying a variety of demographic metrics of interest including race/ethnicity, income, and employment characteristics.

```{r Typology Leaflet Map, echo = FALSE, message=FALSE, warning=FALSE, layout="l-body", fig.height=6}
popup_pa <- paste0("<b>", "GEOID: ", "</b>", priorityareas$GEOID, "<br>",
               "<b>", "Median Household Income: ", "</b>", scales::dollar(priorityareas$mhiE), "<br>",
               "<b>", "Total Population: ", "</b>", scales::comma(round(priorityareas$tot_popE, digits = 0)), "<br>",
               "<b>", "Asian (%): ", "</b>", round(priorityareas$as_per*100, digits =1), "%", "<br>",
               "<b>", "Black (%): ", "</b>", round(priorityareas$bl_per*100, digits =1), "%", "<br>",
               "<b>", "Hispanic (%) : ", "</b>", round(priorityareas$hi_per*100, digits =1), "%", "<br>",
               "<b>", "White (%): ", "</b>", round(priorityareas$wh_per*100, digits =1), "%", "<br>",
               "<b>", "Child Poverty Rate: ", "</b>", round(priorityareas$cpr*100, digits =1), "%", "<br>",
               "<b>", "Internet Access (%): ", "</b>", round(priorityareas$int_per*100, digits =1), "%", "<br>",
               "<b>", "Single-Mom Households (%): ", "</b>", round(priorityareas$smhh_per*100, digits =1), "%", "<br>",
               "<b>", "Single-Mom Households Below Poverty (%): ", "</b>", round(priorityareas$smhh_bpE*100, digits =1), "%", "<br>",
               "<b>", "Unemployment Rate: ", "</b>", round(priorityareas$unemE*100, digits =1), "%"
               )

popup_sd <- paste0("<b>", "GEOID: ", "</b>", sup_dem$GEOID, "<br>",
               "<b>", "Median Household Income: ", "</b>", scales::dollar(sup_dem$mhiE), "<br>",
               "<b>", "Total Population: ", "</b>", scales::comma(round(sup_dem$tot_popE, digits = 0)), "<br>",
               "<b>", "Asian (%): ", "</b>", round(sup_dem$as_per*100, digits =1), "%", "<br>",
               "<b>", "Black (%): ", "</b>", round(sup_dem$bl_per*100, digits =1), "%", "<br>",
               "<b>", "Hispanic (%) : ", "</b>", round(sup_dem$hi_per*100, digits =1), "%", "<br>",
               "<b>", "White (%): ", "</b>", round(sup_dem$wh_per*100, digits =1), "%", "<br>",
               "<b>", "Child Poverty Rate: ", "</b>", round(sup_dem$cpr*100, digits =1), "%", "<br>",
               "<b>", "Internet Access (%): ", "</b>", round(sup_dem$int_per*100, digits =1), "%", "<br>",
               "<b>", "Single-Mom Households (%): ", "</b>", round(sup_dem$smhh_per*100, digits =1), "%", "<br>",
               "<b>", "Single-Mom Households Below Poverty (%): ", "</b>", round(sup_dem$smhh_bpE*100, digits =1), "%", "<br>",
               "<b>", "Unemployment Rate: ", "</b>", round(sup_dem$unemE*100, digits =1), "%"
               )

popup_ds <- paste0("<b>", "GEOID: ", "</b>", dem_sup$GEOID, "<br>",
               "<b>", "Median Household Income: ", "</b>", scales::dollar(dem_sup$mhiE), "<br>",
               "<b>", "Total Population: ", "</b>", scales::comma(round(dem_sup$tot_popE, digits = 0)), "<br>",
               "<b>", "Asian (%): ", "</b>", round(dem_sup$as_per*100, digits =1), "%", "<br>",
               "<b>", "Black (%): ", "</b>", round(dem_sup$bl_per*100, digits =1), "%", "<br>",
               "<b>", "Hispanic (%) : ", "</b>", round(dem_sup$hi_per*100, digits =1), "%", "<br>",
               "<b>", "White (%): ", "</b>", round(dem_sup$wh_per*100, digits =1), "%", "<br>",
               "<b>", "Child Poverty Rate (u18): ", "</b>", round(dem_sup$cpr*100, digits =1), "%", "<br>",
               "<b>", "Internet Access (%): ", "</b>", round(dem_sup$int_per*100, digits =1), "%", "<br>",
               "<b>", "Single-Mom Households (%): ", "</b>", round(dem_sup$smhh_per*100, digits =1), "%", "<br>",
               "<b>", "Single-Mom Households Below Poverty (%): ", "</b>", round(dem_sup$smhh_bpE*100, digits =1), "%", "<br>",
               "<b>", "Unemployment Rate: ", "</b>", round(dem_sup$unemE*100, digits =1), "%"
               )

popup_ti <- paste0("<b>", "GEOID: ", "</b>", demTsup$GEOID, "<br>",
               "<b>", "Median Household Income: ", "</b>", scales::dollar(demTsup$mhiE), "<br>",
               "<b>", "Total Population: ", "</b>", scales::comma(round(demTsup$tot_popE, digits = 0)), "<br>",
               "<b>", "Asian (%): ", "</b>", round(demTsup$as_per*100, digits =1), "%", "<br>",
               "<b>", "Black (%): ", "</b>", round(demTsup$bl_per*100, digits =1), "%", "<br>",
               "<b>", "Hispanic (%) : ", "</b>", round(demTsup$hi_per*100, digits =1), "%", "<br>",
               "<b>", "White (%): ", "</b>", round(demTsup$wh_per*100, digits =1), "%", "<br>",
               "<b>", "Child Poverty Rate: ", "</b>", round(demTsup$cpr*100, digits =1), "%", "<br>",
               "<b>", "Internet Access (%): ", "</b>", round(demTsup$int_per*100, digits =1), "%", "<br>",
               "<b>", "Single-Mom Households (%): ", "</b>", round(demTsup$smhh_per*100, digits =1), "%", "<br>",
               "<b>", "Single-Mom Households Below Poverty (%): ", "</b>", round(demTsup$smhh_bpE*100, digits =1), "%", "<br>",
               "<b>", "Unemployment Rate: ", "</b>", round(demTsup$unemE*100, digits =1), "%"
               )

leaflet() %>%
#  setView(lng = -96.7970, lat = 32.7767, zoom = 10) %>%
  addTiles(urlTemplate = cpal_mapbox, attribution = cpal_leaflet) %>%
    addPolygons(data = priorityareas,
              fillColor = "#84301f",
              weight = 1,
              opacity = 1,
              color = "#84301f ",
              fillOpacity = 0.7,
              popup = ~popup_pa,
              group = "Priority Areas"
              ) %>%
  addPolygons(data = dem_sup,
              fillColor = "#DC5034",
              weight = 1,
              opacity = 1,
              color = "#DC5034",
              fillOpacity = 0.7,
              popup = ~popup_ds,
              group = "Demand Exceeds Supply"
              ) %>%
  addPolygons(data = demTsup,
              fillColor = "#8996A0",
              weight = 1,
              opacity = 1,
              color = "#8996A0",
              fillOpacity = 0.6,
              popup = ~popup_ti,
              group = "Demand Meets Supply"
              ) %>%
  addPolygons(data = sup_dem,
              fillColor = "#0089C4",
              weight = 1,
              opacity = 1,
              color = "#0089C4",
              fillOpacity = 0.6,
              popup = ~popup_sd,
              group = "Supply Exceeds Demand"
              ) %>%
    addPolygons(data = dallasCounty,
              color = "#686158",
              weight = 3,
              smoothFactor = 0.2, 
              fillOpacity = 0,
              opacity = 1,
              group = "Dallas County"
              ) %>%
  addPolygons(data = citycouncil,
              color = "#EDE8D5",
              weight = 3,
              smoothFactor = 0.2, 
              fillOpacity = 0,
              opacity = 1,
              group = "Dallas City Council",
              label = ~htmlEscape(name)
              ) %>%
  addLayersControl(
    baseGroups = c("Dallas County"),
    overlayGroups = c("Dallas City Council", "Priority Areas", "Demand Exceeds Supply", "Supply Exceeds Demand", "Demand Meets Supply"),
    position = "topright",
    options = layersControlOptions(collapsed = FALSE)) %>% 
  hideGroup(c("Demand Exceeds Supply", "Supply Exceeds Demand", "Demand Meets Supply", "Dallas City Council"))
```

## Classification  of Afterschool Program Access

To assess which census tracts have the highest demand compared to supply, each score was classified into one of five initial groups,  ranging from very low to very high.

The group classifications for the demand and supply scores were then compared against one another in order to determine a census tract's general characteristics.

## Primary Classifications 

By comparing the distance between the supply and demand score group against one another we were able to identify three primary classifications  within Dallas County:

* **Supply Exceeds Demand:** A census tract where the supply score significantly exceeds the demand score. Census tracts in this category tend to have a higher number of program seats available relative to the eligible population.

* **Demand Exceeds Supply:** A census tract where the demand score significantly exceeds the supply score. Census tracts in this category tend to have a lower number of program seats available relative to the eligible population.

* **Supply Meets Demand:** A census tract where the supply score and demand score are similar to one another. Census tracts in this category tend to have a similar number of available program seats relative to the eligible population.

```{r, echo = FALSE, message=FALSE, warning=FALSE, layout="l-body-outset", fig.height=8}
typologytable %>%
  filter(!is.na(MainGrp)) %>%
  mutate(MainGrp = str_to_title(MainGrp),
         AreaSqMi = round(AreaSqMi),
         popden = round(eligPop/AreaSqMi),
         mhi = scales::dollar(round(mhi)),
         eligPop = scales::comma(eligPop),
         dasSeats = scales::comma(dasSeats),
         cpr = paste0(round(cpr*100, digits = 1), "%"),
         poppro = round(poppro, digits = 1)) %>%
  select(MainGrp, TotalTracts, AreaSqMi, mhi, eligPop, dasPrograms, dasSeats, poppro, cpr, popden) %>%
  knitr::kable(.,
               col.names = c('Typology', 'Total Tracts', 'Area SqMi', 'Median Household Income', 'Eligible Children', 'Total Programs', 'Total Seats', 'Eligible to Seat Ratio', 'Child Poverty Rate', 'Eligible Density')) %>%
  kable_styling()
```

## Identifying Priority Areas

The group of census tracts identified in the 'Demand Exceeds Supply' category were refined into higher priority communities.

These census tracts contain the the following characteristics:

* Total number of eligible children above the mean of Dallas County (590).

* Household median income  below the mean of Dallas County ($71,739).

* A child poverty rate above the mean for Dallas County (19%).

```{r priority area to non priority table, echo = FALSE, message=FALSE, warning=FALSE, layout="l-body-outset", fig.height=8}
priorityareas %>%
  st_drop_geometry(.) %>%
  rbind(., st_drop_geometry(nopriorityareas)) %>%
  group_by(MainGrp) %>%
  summarise(TotalTracts = n(),
            AreaSqMi = sum(AreaSqMi),
            mhi = median(med_incE, na.rm = TRUE),
            eligPop = sum(eligPop),
            bp_u18 = sum(bp_u18E, na.rm = TRUE),
            pop_u18 = sum(pop_u18E, na.rm = TRUE),
            dasPrograms = sum(dasPrograms),
            dasSeats = sum(dasSeats),
            #cacfpPrograms = sum(cacfpPrograms),
            #cacfpSeats = sum(cacfpSeats),
            #hhsPrograms = sum(hhsPrograms),
            #hhsSeats = sum(hhsSeats),
            #totPrograms = sum(dasPrograms+cacfpPrograms+hhsPrograms),
            #totSeats = sum(dasSeats+cacfpSeats+hhsSeats)
            ) %>%
  mutate(poppro = round(eligPop/dasSeats, digits = 3),
         cpr = round(bp_u18/pop_u18, digits = 3)) %>%
  mutate(MainGrp = str_to_title(MainGrp),
         AreaSqMi = round(AreaSqMi),
         popden = round(eligPop/AreaSqMi),
         mhi = scales::dollar(round(mhi)),
         eligPop = scales::comma(eligPop),
         dasSeats = scales::comma(dasSeats),
         cpr = paste0(round(cpr*100, digits = 1), "%"),
         poppro = round(poppro, digits = 1)) %>%
  select(MainGrp, TotalTracts, AreaSqMi, mhi, eligPop, dasPrograms, dasSeats, poppro, cpr, popden) %>%
  knitr::kable(.,
               col.names = c('Typology', 'Total Tracts', 'Area SqMi', 'Median Household Income', 'Eligible Children', 'Total Programs', 'Total Seats', 'Eligible to Seat Ratio', 'Child Poverty Rate', 'Eligible Density')) %>%
  kable_styling()
```

Forty four census tracts were identified as high priority.

There are 38,400 eligible children living in these 44 census tracts, yet there are no programs available for families. 

In contrast, there are 35 high supply census tracts relative to current demand. These tracts  have an estimated 13,391 eligible children  with a total of 84 afterschool  programs. 

These disparities emphasize the need for a more equitable distribution of afterschool programs in the areas of highest need.. 