---
title: "Dallas County Supply and Demand Scores"
output:
  distill::distill_article:
    toc: TRUE
    theme: theme.css
---

The map below visualizes the supply and demand scores for each census tract. By toggling the layer selection on the top right of the map you will be able to view each score. By clicking on any given census tract a popup will appear displaying a variety of demographic metrics including race/ethnicity, income, and employment characteristics.

```{r, Component Index, echo = FALSE, message=FALSE, warning=FALSE, layout="l-body", fig.height=6}
pal_das <- c("#DC5034", "#0089C4", "#FFB533", "#ADB533", "#8078B6", "#8996A0", "#54494B")
#names(dasDemo)
popup <- paste0("<b>", "GEOID: ", "</b>", dasmetrics$GEOID, "<br>",
                "<b>", "Demand Score: ", "</b>", dasmetrics$indexDemand, "<br>",
                "<b>", "Supply Score: ", "</b>", dasmetrics$indexSupply, "<br>",
               "<b>", "Median Household Income: ", "</b>", scales::dollar(dasmetrics$mhiE), "<br>",
               "<b>", "Total Population: ", "</b>", scales::comma(round(dasmetrics$tot_popE, digits = 0)), "<br>",
               "<b>", "Asian (%): ", "</b>", round(dasmetrics$as_per*100, digits =1), "%", "<br>",
               "<b>", "Black (%): ", "</b>", round(dasmetrics$bl_per*100, digits =1), "%", "<br>",
               "<b>", "Hispanic (%) : ", "</b>", round(dasmetrics$hi_per*100, digits =1), "%", "<br>",
               "<b>", "White (%): ", "</b>", round(dasmetrics$wh_per*100, digits =1), "%", "<br>",
               "<b>", "Child Poverty Rate: ", "</b>", round(dasmetrics$cpr*100, digits =1), "%", "<br>",
               "<b>", "Internet Access (%): ", "</b>", round(dasmetrics$int_per*100, digits =1), "%", "<br>",
               "<b>", "Single-Mom Households (%): ", "</b>", round(dasmetrics$smhh_per*100, digits =1), "%", "<br>",
               "<b>", "Single-Mom Households Below Poverty (%): ", "</b>", round(dasmetrics$smhh_bpE*100, digits =1), "%", "<br>",
               "<b>", "Unemployment Rate: ", "</b>", round(dasmetrics$unemE*100, digits =1), "%"
               )

brksDemand <- BAMMtools::getJenksBreaks(k = 5, var = dasmetrics$indexDemand)
brksSupply <- BAMMtools::getJenksBreaks(k = 5, var = dasmetrics$indexSupply)

supply_1 <- dasmetrics %>%
  filter(indexSupply >= 0 & indexSupply <= 22)
supply_2 <- dasmetrics %>%
  filter(indexSupply > 22 & indexSupply <= 56)
supply_3 <- dasmetrics %>%
  filter(indexSupply > 56 & indexSupply <= 84)
supply_4 <- dasmetrics %>%
  filter(indexSupply > 84 & indexSupply <= 100)
supply_5 <- dasmetrics %>%
  filter(is.na(indexSupply))

demand_1 <- dasmetrics %>%
  filter(indexDemand >= 0 & indexDemand <= 42)
demand_2 <- dasmetrics %>%
  filter(indexDemand > 42 & indexDemand <= 66)
demand_3 <- dasmetrics %>%
  filter(indexDemand > 66 & indexDemand <= 79)
demand_4 <- dasmetrics %>%
  filter(indexDemand > 79 & indexDemand <= 100)
demand_5 <- dasmetrics %>%
  filter(is.na(indexDemand))

palSupply <- colorBin("YlOrBr", domain = dasComponents$indexSupply, bins = brksSupply)
palDemand <- colorBin("BuPu", domain = dasComponents$indexDemand, bins = brksDemand)

leaflet() %>%
#  setView(lng = -96.7970, lat = 32.7767, zoom = 10) %>%
  addTiles(urlTemplate = cpal_mapbox, attribution = cpal_leaflet) %>%
  addPolygons(data = demand_1,
              fillColor = "#edf8fb",
              color = "#595959",
              weight = 0.5,
              opacity = 0.8,
              fillOpacity = 0.4,
              group = "Demand Very Low",
              popup = ~popup,
              highlightOptions = highlightOptions(color = "white", 
                                                  weight = 2,
                                                  bringToFront = TRUE)
              ) %>%
  addPolygons(data = demand_2,
              fillColor = "#b3cde3",
              color = "#595959",
              weight = 0.5,
              opacity = 0.8,
              fillOpacity = 0.4,
              group = "Demand Low",
              popup = ~popup,
              highlightOptions = highlightOptions(color = "white", 
                                                  weight = 2,
                                                  bringToFront = TRUE)
              ) %>%
  addPolygons(data = demand_3,
              fillColor = "#8c96c6",
              color = "#595959",
              weight = 0.5,
              opacity = 0.8,
              fillOpacity = 0.4,
              group = "Demand High",
              popup = ~popup,
              highlightOptions = highlightOptions(color = "white", 
                                                  weight = 2,
                                                  bringToFront = TRUE)
              ) %>%
  addPolygons(data = demand_4,
              fillColor = "#88419d",
              color = "#595959",
              weight = 0.5,
              opacity = 0.8,
              fillOpacity = 0.4,
              group = "Demand Very High",
              popup = ~popup,
              highlightOptions = highlightOptions(color = "white", 
                                                  weight = 2,
                                                  bringToFront = TRUE)
              ) %>%
  addPolygons(data = demand_5,
              fillColor = "#595959",
              color = "#595959",
              weight = 0.5,
              opacity = 0.8,
              fillOpacity = 0.4,
              group = "Demand Unavailable",
              popup = ~popup,
              highlightOptions = highlightOptions(color = "white", 
                                                  weight = 2,
                                                  bringToFront = TRUE)
              ) %>%
    addPolygons(data = supply_1,
              fillColor = "#ffffd4",
              color = "#595959",
              weight = 0.5,
              opacity = 0.8,
              fillOpacity = 0.4,
              group = "Supply Very Low",
              popup = ~popup,
              highlightOptions = highlightOptions(color = "white", 
                                                  weight = 2,
                                                  bringToFront = TRUE)
              ) %>%
    addPolygons(data = supply_2,
              fillColor = "#fed98e",
              color = "#595959",
              weight = 0.5,
              opacity = 0.8,
              fillOpacity = 0.4,
              group = "Supply Low",
              popup = ~popup,
              highlightOptions = highlightOptions(color = "white", 
                                                  weight = 2,
                                                  bringToFront = TRUE)
              ) %>%
    addPolygons(data = supply_3,
              fillColor = "#fe9929",
              color = "#595959",
              weight = 0.5,
              opacity = 0.8,
              fillOpacity = 0.4,
              group = "Supply High",
              popup = ~popup,
              highlightOptions = highlightOptions(color = "white", 
                                                  weight = 2,
                                                  bringToFront = TRUE)
              ) %>%
    addPolygons(data = supply_4,
              fillColor = "#cc4c02",
              color = "#595959",
              weight = 0.5,
              opacity = 0.8,
              fillOpacity = 0.4,
              group = "Supply Very High",
              popup = ~popup,
              highlightOptions = highlightOptions(color = "white", 
                                                  weight = 2,
                                                  bringToFront = TRUE)
              ) %>%
    addPolygons(data = supply_5,
              fillColor = "#595959",
              color = "#595959",
              weight = 0.5,
              opacity = 0.8,
              fillOpacity = 0.4,
              group = "Supply Unavailable",
              popup = ~popup,
              highlightOptions = highlightOptions(color = "white", 
                                                  weight = 2,
                                                  bringToFront = TRUE)
              ) %>%
  addLegend(data = dasmetrics,
            "bottomleft", 
            pal = palDemand, 
            values = ~indexDemand,
            title = "Demand Score",
            group = "Demand Very Low",
            opacity = 1) %>%
  addLegend(data = dasmetrics,
            "bottomleft", 
            pal = palDemand, 
            values = ~indexDemand,
            title = "Demand Score",
            group = "Demand Low",
            opacity = 1) %>%
  addLegend(data = dasmetrics,
            "bottomleft", 
            pal = palDemand, 
            values = ~indexDemand,
            title = "Demand Score",
            group = "Demand High",
            opacity = 1) %>%
  addLegend(data = dasmetrics,
            "bottomleft", 
            pal = palDemand, 
            values = ~indexDemand,
            title = "Demand Score",
            group = "Demand Very High",
            opacity = 1) %>%
  addLegend(data = dasmetrics,
            "bottomleft", 
            pal = palDemand, 
            values = ~indexDemand,
            title = "Demand Score",
            group = "Demand Unavailable",
            opacity = 1) %>%
  addLayersControl(
    overlayGroups = c("Demand Very Low",
                      "Demand Low",
                      "Demand High",
                      "Demand Very High",
                      "Demand Unavailable",
                      "Supply Very Low",
                      "Supply Low",
                      "Supply High",
                      "Supply Very High",
                      "Supply Unavailable"),
    position = "topright",
    options = layersControlOptions(collapsed = FALSE)) %>% 
  hideGroup(c("Supply Very Low",
              "Supply Low",
              "Supply High",
              "Supply Very High",
              "Supply Unavailable")) %>%
  htmlwidgets::onRender("
    function(el, x) {
      var updateLegend = function () {
          var selectedGroup = document.querySelectorAll('input:checked')[0].nextSibling.innerText.substr(1);

          document.querySelectorAll('.legend').forEach(a => a.hidden=true);
          document.querySelectorAll('.legend').forEach(l => {
            if (l.children[0].children[0].innerText == selectedGroup) l.hidden=false;
          });
      };
      updateLegend();
      this.on('baselayerchange', e => updateLegend());
    }")
```

## Supply and Demand in Dallas County

The demand for afterschool programming is consistently high across Dallas County. However, the supply of afterschool programs is  concentrated in certain communities. About half of all census tracts in Dallas County received a score of 75 or greater on supply. These 379 census tracts contain **97%** of all afterschool programs in the county. Yet, only **63%** of eligible children live in these tracts. 

```{r, Supply Scores Histogram, echo = FALSE, message=FALSE, warning=FALSE, include=FALSE}
dasmetrics %>%
  st_drop_geometry(.) %>%
  filter(indexSupply >= 75) %>%
  summarize(count = n(),
            programs = sum(dasPrograms),
            seats = sum(dasSeats),
            eligpop = sum(eligPop))

857/903

240602/sum(dasmetrics$eligPop)

dasComponents %>%
  ggplot(aes(x = indexSupply)) +
  geom_histogram(bins = 50, color = "#0089C4", fill = "#0089C4", alpha = 0.7) +
  labs(title = "Distribution of Supply Scores",
       #subtitle = "U.S. Census Bureau, American Community Survey 5-Year Estimates 2020",
       x = "Supply Score",
       y = "Total Census Tracts",
       color = "") +
  theme_bw() +
  theme(legend.position = "bottom",
        legend.key.width = unit(0.5, units = "in"),
        plot.caption = element_text(size = 36, face = "italic"),
        legend.spacing = unit(0, "in"), 
        legend.key.height = unit(0.25, "in"), 
        legend.spacing.x = unit(0.1, "in"), 
        legend.justification = "center",
        legend.margin=margin(4,4,4,4),
        legend.box.spacing = margin(0.5))
```


```{r, Demand Scores Histogram, echo = FALSE, message=FALSE, warning=FALSE, include=FALSE}
dasComponents %>%
  ggplot(aes(x = indexDemand)) +
  geom_histogram(bins = 50, color = "#DC5034", fill = "#DC5034", alpha = 0.7) +
  labs(title = "Distribution of Demand Scores",
       #subtitle = "U.S. Census Bureau, American Community Survey 5-Year Estimates 2020",
       x = "Demand Score",
       y = "Total Census Tracts",
       color = "") +
  theme_bw() +
  theme(legend.position = "bottom",
        legend.key.width = unit(0.5, units = "in"),
        plot.caption = element_text(size = 36, face = "italic"),
        legend.spacing = unit(0, "in"), 
        legend.key.height = unit(0.25, "in"), 
        legend.spacing.x = unit(0.1, "in"), 
        legend.justification = "center",
        legend.margin=margin(4,4,4,4),
        legend.box.spacing = margin(0.5))

dasmetrics %>%
  st_drop_geometry(.) %>%
  filter(indexDemand <= 50) %>%
  summarize(count = n(),
            programs = sum(dasPrograms),
            seats = sum(dasSeats),
            eligpop = sum(eligPop))
```

## What do the Scores Mean?

The scores are not meant to identify the exact number of seats in a census tract. The afterschool environment is dynamic. Programs may open or close, change locations, or add or remove seats based on changes in staffing, funding, or other external factors. Dallas Afterschool monitors the overall afterschool landscape, but specific changes are not always available in real time. Additionally, the characteristics of local communities, measured by the U.S. Census Bureau, are based on a relatively small - but robust - sample of the population. Given the changing nature of the afterschool environment, external factors such as the impact of COVID, may not be reflected in the underlying census data.

As a result, our intention is not to identify the exact number of afterschool seats needed to address unmet demand. For example, data could indicate that a program has capacity for 50 students, but it will not show if a program has a waitlist. To account for this gap, we developed two sets of scores to measure the relative position of a neighborhood's afterschool environment: 1) the supply of seats and programs and 2) the potential need or demand for afterschool programming based on population characteristics. 

We assessed these two components through the creation of two indices, which produced scores ranging from 0 to 100. A census tract with a Supply Score of 58 suggests that this particular location falls in the middle when it comes to the availability of afterschool programs and seats.

